---
title: "Bayesian Hierarchical Model for PM2.5"
author: "Leila Cielok"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
set.seed(123)
```

# 1. Introduction

This project implements a Bayesian hierarchical model for daily PM2.5 concentrations observed across multiple monitoring stations. The outcome is modeled on the log scale to account for right-skewness in the raw data.

The specification is:

\[
\tilde y_i = \log(y_i), \qquad
\tilde y_i \mid \beta, \alpha, \sigma^2 \sim \mathcal N(x_i'\beta + \alpha_{j[i]}, \sigma^2)
\]

with station-specific random intercepts

\[
\alpha_j \mid \tau \sim \mathcal N(0, \tau^2), \qquad j = 1, \dots, J.
\]

A Metropolis-within-Gibbs sampler is used:
- Gibbs updates for \(\beta\), \(\alpha_j\), and \(\sigma^2\)
- a Metropolis-Hastings step for \(\tau\)

# 2. Libraries and data import

```{r libraries}
library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)
library(stringr)
library(purrr)
```

```{r load-data}
df <- read_csv("../pm25_bayes.csv", show_col_types = FALSE)
str(df)
df$Pollutant <- NULL
summary(df)
```

# 3. Data preprocessing

```{r preprocessing}
# If log_y is not already present, create it from y or Value
if (!"log_y" %in% names(df)) {
  if ("y" %in% names(df)) {
    df <- df %>% mutate(log_y = log(y))
  } else if ("Value" %in% names(df)) {
    df <- df %>% mutate(log_y = log(Value))
  } else {
    stop("Neither log_y nor y/Value is available in the dataset.")
  }
}

# Convert categorical variables
# Keep explicit reference levels if you want stable interpretation
df <- df %>%
  mutate(
    station_id = factor(station_id),
    month = factor(month),
    day_of_week = factor(day_of_week)
  )

# Optional: reorder levels if needed
# df$month <- factor(df$month, levels = c("1","2","3",...))
# df$day_of_week <- factor(df$day_of_week,
#                          levels = c("Monday","Tuesday","Wednesday","Thursday",
#                                     "Friday","Saturday","Sunday"))

# Keep only complete cases for the model variables
model_vars <- c("log_y", "station_id", "month", "day_of_week")
df_model <- df %>%
  select(all_of(model_vars)) %>%
  drop_na()

cat("Number of observations:", nrow(df_model), "\n")
cat("Number of stations:", n_distinct(df_model$station_id), "\n")
```

## 3.1 Optional standardization of continuous covariates

In the current specification, `month` and `day_of_week` are categorical, so there is nothing to standardize among the regressors. If later you add continuous covariates, standardize only those columns, not the outcome and not dummy variables.

```{r standardization-helper}
standardize_vars <- function(data, vars) {
  data %>%
    mutate(across(all_of(vars), ~ as.numeric(scale(.x)), .names = "{.col}_std"))
}

# Example (ONLY if you later add continuous regressors):
# df_model <- standardize_vars(df_model, c("temperature", "humidity"))
```

# 4. Exploratory analysis

```{r eda-plots, fig.width=10, fig.height=4}
p1 <- ggplot(df_model, aes(log_y)) +
  geom_histogram(bins = 50) +
  labs(title = "Distribution of log(PM2.5)")

# If raw y exists in the original dataset, compare with log_y
p2 <- NULL
if ("y" %in% names(df)) {
  p2 <- ggplot(df, aes(y)) +
    geom_histogram(bins = 50) +
    labs(title = "Distribution of raw PM2.5")
} else if ("Value" %in% names(df)) {
  p2 <- ggplot(df, aes(Value)) +
    geom_histogram(bins = 50) +
    labs(title = "Distribution of raw PM2.5")
}

print(p1)
if (!is.null(p2)) print(p2)
```

```{r eda-summary}
df_model %>%
  summarise(
    n = n(),
    mean_log_y = mean(log_y),
    sd_log_y = sd(log_y),
    min_log_y = min(log_y),
    max_log_y = max(log_y)
  )
```

# 5. Design matrices and indexing

```{r design-matrices}
# Fixed-effects matrix
X <- model.matrix(~ month + day_of_week, data = df_model)

# Outcome vector
y_tilde <- df_model$log_y

# Group index for stations
station_index <- as.integer(df_model$station_id)
J <- nlevels(df_model$station_id)
n <- length(y_tilde)
p <- ncol(X)

# Number of observations per group
n_j <- as.integer(table(df_model$station_id))

cat("n =", n, "\n")
cat("p =", p, "\n")
cat("J =", J, "\n")
```

# 6. Prior specification

```{r priors}
# Prior for beta ~ N(b0, B0)
# Baseline choice; change c_prior for sensitivity analysis
c_prior <- 100
b0 <- rep(0, p)
B0 <- diag(c_prior, p)
B0_inv <- solve(B0)

# Prior for sigma^2 ~ IG(a_sigma, b_sigma)
a_sigma <- 2
b_sigma <- 0.5   # can compare with 1 in sensitivity analysis

# Prior for tau ~ Half-Cauchy(0, A)
A_tau <- 5       # can compare with 2.5 and 10 in sensitivity analysis

list(c_prior = c_prior, a_sigma = a_sigma, b_sigma = b_sigma, A_tau = A_tau)
```

# 7. Full conditional formulas used in the sampler

The sampler uses the following updates:

\[
\beta \mid \cdot \sim \mathcal N(m_\beta, V_\beta)
\]

with

\[
V_\beta = \left(B_0^{-1} + \frac{1}{\sigma^2}X'X\right)^{-1},
\qquad
m_\beta = V_\beta\left(B_0^{-1}b_0 + \frac{1}{\sigma^2}X'(\tilde y - Z\alpha)\right)
\]

\[
\alpha_j \mid \cdot \sim \mathcal N(m_j, V_j)
\]

with

\[
V_j = \left(\frac{n_j}{\sigma^2} + \frac{1}{\tau^2}\right)^{-1},
\qquad
m_j = V_j\frac{1}{\sigma^2}\sum_{i \in j}(\tilde y_i - x_i'\beta)
\]

\[
\sigma^2 \mid \cdot \sim IG\left(a_\sigma + \frac{n}{2}, b_\sigma + \frac{SSR}{2}\right)
\]

For \(\tau\), the full conditional is only known up to proportionality, so a Metropolis-Hastings step is used.

# 8. Utility functions for the sampler

```{r utility-functions}
# Draw from Inverse-Gamma(shape, scale) under the convention:
# p(x) propto x^{-(a+1)} exp(-b/x)
# If G ~ Gamma(shape=a, rate=b), then 1/G ~ IG(a,b)
rinvgamma <- function(n, shape, scale) {
  1 / rgamma(n, shape = shape, rate = scale)
}

# Log-target for eta = log(tau)
log_target_eta <- function(eta, alpha, A_tau) {
  J <- length(alpha)
  S_alpha <- sum(alpha^2)
  -(J - 1) * eta - S_alpha / (2 * exp(2 * eta)) - log(1 + exp(2 * eta) / A_tau^2)
}
```

# 9. Metropolis-within-Gibbs sampler

```{r sampler-function}
run_mwg_sampler <- function(y_tilde, X, station_index, n_j,
                            b0, B0, a_sigma, b_sigma, A_tau,
                            n_iter = 5000, burn_in = 1000,
                            proposal_sd = 0.15,
                            seed = 123) {

  set.seed(seed)

  n <- length(y_tilde)
  p <- ncol(X)
  J <- length(n_j)
  B0_inv <- solve(B0)

  # Storage
  beta_store <- matrix(NA_real_, nrow = n_iter, ncol = p)
  alpha_store <- matrix(NA_real_, nrow = n_iter, ncol = J)
  sigma2_store <- numeric(n_iter)
  tau_store <- numeric(n_iter)
  accept_tau <- logical(n_iter)

  # Initial values
  beta_curr <- rep(0, p)
  alpha_curr <- rep(0, J)
  sigma2_curr <- var(y_tilde)
  tau_curr <- 1

  XtX <- crossprod(X)

  for (t in seq_len(n_iter)) {

    # ---- 1. Gibbs update for beta ----
    alpha_vec <- alpha_curr[station_index]
    V_beta <- solve(B0_inv + XtX / sigma2_curr)
    m_beta <- V_beta %*% (B0_inv %*% b0 + crossprod(X, y_tilde - alpha_vec) / sigma2_curr)
    beta_curr <- as.numeric(MASS::mvrnorm(1, mu = as.numeric(m_beta), Sigma = V_beta))

    # ---- 2. Gibbs updates for alpha_j ----
    xb <- as.numeric(X %*% beta_curr)

    for (j in seq_len(J)) {
      idx <- which(station_index == j)
      V_j <- 1 / (length(idx) / sigma2_curr + 1 / (tau_curr^2))
      m_j <- V_j * sum(y_tilde[idx] - xb[idx]) / sigma2_curr
      alpha_curr[j] <- rnorm(1, mean = m_j, sd = sqrt(V_j))
    }

    # ---- 3. Gibbs update for sigma^2 ----
    resid <- y_tilde - xb - alpha_curr[station_index]
    SSR <- sum(resid^2)
    sigma2_curr <- rinvgamma(1, shape = a_sigma + n / 2, scale = b_sigma + SSR / 2)

    # ---- 4. MH update for tau on log scale ----
    eta_curr <- log(tau_curr)
    eta_prop <- rnorm(1, mean = eta_curr, sd = proposal_sd)

    log_r <- log_target_eta(eta_prop, alpha_curr, A_tau) -
      log_target_eta(eta_curr, alpha_curr, A_tau)

    if (log(runif(1)) < log_r) {
      tau_curr <- exp(eta_prop)
      accept_tau[t] <- TRUE
    } else {
      accept_tau[t] <- FALSE
    }

    # ---- Store current draws ----
    beta_store[t, ] <- beta_curr
    alpha_store[t, ] <- alpha_curr
    sigma2_store[t] <- sigma2_curr
    tau_store[t] <- tau_curr
  }

  kept <- seq.int(from = burn_in + 1, to = n_iter)

  list(
    beta = beta_store,
    alpha = alpha_store,
    sigma2 = sigma2_store,
    tau = tau_store,
    accept_tau = accept_tau,
    post_beta = beta_store[kept, , drop = FALSE],
    post_alpha = alpha_store[kept, , drop = FALSE],
    post_sigma2 = sigma2_store[kept],
    post_tau = tau_store[kept],
    accept_rate_tau = mean(accept_tau[kept])
  )
}
```

# 10. Run the sampler

```{r run-sampler}
# MASS::mvrnorm is used inside the function
library(MASS)

mcmc_fit <- run_mwg_sampler(
  y_tilde = y_tilde,
  X = X,
  station_index = station_index,
  n_j = n_j,
  b0 = b0,
  B0 = B0,
  a_sigma = a_sigma,
  b_sigma = b_sigma,
  A_tau = A_tau,
  n_iter = 6000,
  burn_in = 1000,
  proposal_sd = 0.12,
  seed = 123
)

mcmc_fit$accept_rate_tau
```

# 11. Posterior summaries

```{r posterior-summaries}
beta_names <- colnames(X)

beta_summary <- data.frame(
  parameter = beta_names,
  mean = apply(mcmc_fit$post_beta, 2, mean),
  sd = apply(mcmc_fit$post_beta, 2, sd),
  q2.5 = apply(mcmc_fit$post_beta, 2, quantile, probs = 0.025),
  q50 = apply(mcmc_fit$post_beta, 2, quantile, probs = 0.50),
  q97.5 = apply(mcmc_fit$post_beta, 2, quantile, probs = 0.975)
)

sigma2_summary <- data.frame(
  parameter = "sigma2",
  mean = mean(mcmc_fit$post_sigma2),
  sd = sd(mcmc_fit$post_sigma2),
  q2.5 = quantile(mcmc_fit$post_sigma2, 0.025),
  q50 = quantile(mcmc_fit$post_sigma2, 0.50),
  q97.5 = quantile(mcmc_fit$post_sigma2, 0.975)
)

tau_summary <- data.frame(
  parameter = "tau",
  mean = mean(mcmc_fit$post_tau),
  sd = sd(mcmc_fit$post_tau),
  q2.5 = quantile(mcmc_fit$post_tau, 0.025),
  q50 = quantile(mcmc_fit$post_tau, 0.50),
  q97.5 = quantile(mcmc_fit$post_tau, 0.975)
)

beta_summary
sigma2_summary
tau_summary
```

# 12. Basic diagnostics

```{r diagnostics, fig.width=10, fig.height=8}
par(mfrow = c(3, 2))
plot(mcmc_fit$post_beta[, 1], type = "l", main = beta_names[1], ylab = "")
if (ncol(mcmc_fit$post_beta) >= 2) plot(mcmc_fit$post_beta[, 2], type = "l", main = beta_names[2], ylab = "")
plot(mcmc_fit$post_sigma2, type = "l", main = expression(sigma^2), ylab = "")
plot(mcmc_fit$post_tau, type = "l", main = expression(tau), ylab = "")
hist(mcmc_fit$post_sigma2, breaks = 40, main = expression(posterior~sigma^2), xlab = expression(sigma^2))
hist(mcmc_fit$post_tau, breaks = 40, main = expression(posterior~tau), xlab = expression(tau))
```

# 13. Sensitivity analysis skeleton

This section can be used to compare prior choices such as:
- \(c \in \{10, 100, 1000\}\) for \(B_0 = cI\)
- \(A \in \{2.5, 5, 10\}\) for the half-Cauchy prior on \(\tau\)
- \(b_\sigma \in \{0.5, 1\}\) for the inverse-gamma prior on \(\sigma^2\)

```{r sensitivity-template}
run_one_spec <- function(c_prior, A_tau, b_sigma,
                         n_iter = 4000, burn_in = 1000) {
  fit <- run_mwg_sampler(
    y_tilde = y_tilde,
    X = X,
    station_index = station_index,
    n_j = n_j,
    b0 = rep(0, ncol(X)),
    B0 = diag(c_prior, ncol(X)),
    a_sigma = 2,
    b_sigma = b_sigma,
    A_tau = A_tau,
    n_iter = n_iter,
    burn_in = burn_in,
    proposal_sd = 0.12,
    seed = 123
  )

  data.frame(
    c_prior = c_prior,
    A_tau = A_tau,
    b_sigma = b_sigma,
    beta_intercept_mean = mean(fit$post_beta[, 1]),
    sigma2_mean = mean(fit$post_sigma2),
    tau_mean = mean(fit$post_tau),
    tau_accept = fit$accept_rate_tau
  )
}

# Example runs (uncomment when ready)
# sensitivity_results <- bind_rows(
#   run_one_spec(c_prior = 10,  A_tau = 2.5, b_sigma = 0.5),
#   run_one_spec(c_prior = 100, A_tau = 5.0, b_sigma = 0.5),
#   run_one_spec(c_prior = 1000, A_tau = 10.0, b_sigma = 1.0)
# )
# sensitivity_results
```

# 14. Notes before final estimation

Before running the final version of the model, check:

1. the acceptance rate for the MH step on \(\tau\) is reasonable;
2. trace plots show stable mixing after burn-in;
3. posterior summaries are stable across different prior specifications;
4. the proposal standard deviation for \(\tau\) is tuned if necessary.

A practical target for the MH acceptance rate is often around 20% to 50%, though there is no single universal threshold.
