---
title: "Bayesian Hierarchical Model for PM2.5"
author: "Nome Cognome"
date: "`r Sys.Date()`"
output: html_document
---

# 1. Introduction

This project implements a Bayesian hierarchical model to analyze daily PM2.5 concentrations across multiple monitoring stations.

The objective is to account for both:
- within-station variability (daily fluctuations)
- between-station heterogeneity


# 2. Data and Preprocessing

```{r}
# Load libraries
library(brms)
library(ggplot2)

# Load data
df <- read.csv("../pm25_bayes.csv")

# Convert categorical variables
df$station_id  <- factor(df$station_id)
df$month       <- factor(df$month)
df$day_of_week <- factor(df$day_of_week)

str(df)
summary(df)
```

# 3. Exploratory Data Analysis

```{r}
par(mfrow=c(1,2))

hist(df$y, breaks=50, main="Distribution of y", col="lightblue")
hist(df$log_y, breaks=50, main="Distribution of log(y)", col="lightgreen")

```
The raw PM2.5 distribution is highly right-skewed, while the log-transformed variable appears approximately symmetric.

```{r}
# Observations per station
obs_per_station <- table(df$station_id)
summary(obs_per_station)

hist(obs_per_station, breaks=30, main="Observations per station", col="lightgray")
```

```{r}
# Variability Analysis
within_var <- tapply(df$log_y, df$station_id, var)
station_means <- tapply(df$log_y, df$station_id, mean)

between_var <- var(station_means)

mean(within_var)
between_var
```
Within-station variability dominates, but between-station variation is non-negligible, motivating a hierarchical structure.

# Model Estimation
```{r}
fit <- brm(
  formula = log_y ~ month + day_of_week + (1 | station_id),
  data = df,
  family = gaussian(),
  prior = c(
    prior(normal(0, 5), class = "Intercept"),
    prior(normal(0, 1), class = "b"),
    prior(student_t(3, 0, 1), class = "sd"),
    prior(student_t(3, 0, 1), class = "sigma")
  ),
  chains = 4,
  iter = 4000,
  warmup = 2000,
  cores = 4,
  seed = 123
)
```
```{r}
summary(fit)
plot(fit)
pp_check(fit)
```
```{r}
saveRDS(fit, "fit_pm25.rds")
```

The model reveals strong seasonal patterns in PM2.5 concentrations, with several months exhibiting substantially lower pollution levels compared to the reference month. In contrast, day-of-week effects are small and mostly not statistically significant.

The estimated standard deviation of station-level random effects indicates meaningful heterogeneity across monitoring stations, confirming the appropriateness of a hierarchical specification.

Posterior predictive checks show that the model adequately captures the distribution of the observed data.

The model reveals strong seasonal patterns in PM2.5 concentrations, with several months exhibiting substantially lower pollution levels compared to the reference month. In contrast, day-of-week effects are small and mostly not statistically significant.

The estimated standard deviation of station-level random effects indicates meaningful heterogeneity across monitoring stations, confirming the appropriateness of a hierarchical specification.

Posterior predictive checks show that the model adequately captures the distribution of the observed data.
