---
title: "Bayesian Hierarchical Model for PM2.5"
author: "Leila Cielok"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
set.seed(123)
```

# Introduction

This project implements a Bayesian hierarchical model for daily PM2.5 concentrations observed across multiple monitoring stations. The outcome is modeled on the log scale to account for right-skewness in the raw data.

The specification is:

\[
\tilde y_i = \log(y_i), \qquad
\tilde y_i \mid \beta, \alpha, \sigma^2 \sim \mathcal N(x_i'\beta + \alpha_{j[i]}, \sigma^2)
\]

with station-specific random intercepts

\[
\alpha_j \mid \tau \sim \mathcal N(0, \tau^2), \qquad j = 1, \dots, J.
\]

A Metropolis-within-Gibbs sampler is used:
- Gibbs updates for \(\beta\), \(\alpha_j\), and \(\sigma^2\)
- a Metropolis-Hastings step for \(\tau\)

# 1. Libraries and data import

```{r libraries}
library(dplyr)
library(ggplot2)
library(readr)
library(e1071)
library(tidyr)
library(stringr)
library(purrr)
```

```{r load-data}
df <- read_csv("../pm25_bayes.csv", show_col_types = FALSE)
str(df)
df$Pollutant <- NULL
df <- df %>%
  mutate(
    Start = as.Date(Start),
    month = factor(month),
    day_of_week = factor(
      day_of_week,
      levels = 0:6,
      labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")
    )
  )
summary(df)
```
# 2 EDA
## 2.1: outcome variable
```{r eda-outcome, message=FALSE, warning=FALSE}
# Basic summary statistics
outcome_var <- "y"

summary(df[[outcome_var]])

# Missing and non-positive values
cat("Number of missing values:", sum(is.na(df[[outcome_var]])), "\n")
cat("Number of non-positive values:", sum(df[[outcome_var]] <= 0, na.rm = TRUE), "\n")

# Histogram of raw outcome
ggplot(df, aes(x = .data[[outcome_var]])) +
  geom_histogram(bins = 40) +
  labs(
    title = "Distribution of the raw PM2.5 outcome",
    x = outcome_var,
    y = "Frequency"
  )

# Skeweness of raw data
skew_raw <- skewness(df[[outcome_var]], na.rm = TRUE)
cat("Skewness (raw):", skew_raw, "\n")

# Histogram of log-transformed outcome
df_pos <- df %>% filter(.data[[outcome_var]] > 0)

ggplot(df_pos, aes(x = log(.data[[outcome_var]]))) +
  geom_histogram(bins = 40) +
  labs(
    title = "Distribution of the log-transformed PM2.5 outcome",
    x = paste0("log(", outcome_var, ")"),
    y = "Frequency"
  )

# Skeweness of log data
skew_log <- skewness(log(df[[outcome_var]]), na.rm = TRUE)
cat("Skewness (log):", skew_log, "\n")

# Compare raw and log-scale summaries
cat("\nSummary of raw outcome:\n")
print(summary(df_pos[[outcome_var]]))

cat("\nSummary of log-transformed outcome:\n")
print(summary(log(df_pos[[outcome_var]])))
```

## 2.2: data
```{r eda-panel-structure, message=FALSE, warning=FALSE}
# 1. Observations per station
obs_per_station <- df %>%
  count(station_id, name = "n_obs") %>%
  arrange(desc(n_obs))

cat("Number of stations:", n_distinct(df$station_id), "\n")
cat("Min observations per station:", min(obs_per_station$n_obs), "\n")
cat("Median observations per station:", median(obs_per_station$n_obs), "\n")
cat("Max observations per station:", max(obs_per_station$n_obs), "\n")

head(obs_per_station, 10)

ggplot(obs_per_station, aes(x = n_obs)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of observations per station",
    x = "Number of observations",
    y = "Frequency"
  )

# 2a Observations per month
obs_per_month <- df %>%
  count(month, name = "n_obs") %>%
  arrange(month)

print(obs_per_month)

ggplot(obs_per_month, aes(x = factor(month), y = n_obs)) +
  geom_col() +
  labs(
    title = "Number of observations by month",
    x = "Month",
    y = "Number of observations"
  )

# 2b Observations per day of week
obs_per_day <- df %>%
  count(day_of_week, name = "n_obs")

print(obs_per_day)

ggplot(obs_per_day, aes(x = day_of_week, y = n_obs)) +
  geom_col() +
  labs(
    title = "Number of observations by day of week",
    x = "Day of week",
    y = "Number of observations"
  )

# 3. Station-month panel structure
station_month <- df %>%
  distinct(station_id, month) %>%
  count(station_id, name = "n_months_present") %>%
  arrange(n_months_present, station_id)

cat("Min number of months observed per station:", min(station_month$n_months_present), "\n")
cat("Median number of months observed per station:", median(station_month$n_months_present), "\n")
cat("Max number of months observed per station:", max(station_month$n_months_present), "\n")

head(station_month, 10)

# 4. Full station x month table and missing months
station_month_full <- expand_grid(
  station_id = sort(unique(df$station_id)),
  month = sort(unique(df$month))
) %>%
  left_join(
    df %>%
      count(station_id, month, name = "n_obs"),
    by = c("station_id", "month")
  ) %>%
  mutate(
    n_obs = ifelse(is.na(n_obs), 0, n_obs),
    missing_month = ifelse(n_obs == 0, 1, 0)
  )

missing_months_by_station <- station_month_full %>%
  group_by(station_id) %>%
  summarise(
    missing_months = sum(missing_month),
    months_observed = sum(n_obs > 0),
    .groups = "drop"
  ) %>%
  arrange(desc(missing_months), station_id)

cat("Stations with at least one completely missing month:",
    sum(missing_months_by_station$missing_months > 0), "\n")

head(missing_months_by_station, 15)

# Optional: show which months are missing for each station
missing_month_detail <- station_month_full %>%
  filter(n_obs == 0) %>%
  arrange(station_id, month)

head(missing_month_detail, 20)

# Heatmap of data availability by station and month
ggplot(station_month_full, aes(x = factor(month), y = factor(station_id), fill = n_obs)) +
  geom_tile() +
  labs(
    title = "Data availability by station and month",
    x = "Month",
    y = "Station ID",
    fill = "N obs"
  ) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```
Exploratory analysis shows that the raw PM2.5 outcome is highly right-skewed, with a skewness of 17.06 and a long upper tail driven by extreme observations. After log transformation, skewness drops to -0.29, yielding a much more symmetric distribution and supporting the use of a Gaussian likelihood for the transformed outcome.

The panel structure is markedly unbalanced across monitoring stations. The number of observations per station ranges from 69 to 332, with a median of 299.5. The dataset is also uneven across months: while most months contain between roughly 2,600 and 3,100 observations, January is severely underrepresented, with only 177 observations. Moreover, 32 stations exhibit at least one completely missing month, and the minimum temporal coverage is only 4 months for some stations. These features support the use of a hierarchical specification, which is well suited to unbalanced grouped data.

# 3. Data preprocessing

```{r preprocessing}
# Create log outcome and keep only finite values
df <- df %>%
  mutate(log_y = log(.data[[outcome_var]])) %>%
  filter(is.finite(log_y))

# Keep only variables used in the model and clean them
model_vars <- c("log_y", "station_id", "month", "day_of_week")

df_model <- df %>%
  dplyr::select(all_of(model_vars)) %>%
  drop_na() %>%
  mutate(
    station_id = factor(station_id),
    month = factor(month),
    day_of_week = factor(day_of_week)
  )

cat("Number of observations:", nrow(df_model), "\n")
cat("Number of stations:", n_distinct(df_model$station_id), "\n")

# Summary of transformed outcome
df_model %>%
  summarise(
    n = n(),
    mean_log_y = mean(log_y),
    sd_log_y = sd(log_y),
    min_log_y = min(log_y),
    max_log_y = max(log_y)
  )
```

# 5. Design matrices and indexing

```{r design-matrices}
# Fixed-effects matrix
X <- model.matrix(~ month + day_of_week, data = df_model)

# Outcome vector
y_tilde <- df_model$log_y

# Group index for stations
station_index <- as.integer(df_model$station_id)

# Dimensions
J <- nlevels(df_model$station_id)
n <- length(y_tilde)
p <- ncol(X)

# Number of observations per group
n_j <- as.integer(table(df_model$station_id))

cat("n =", n, "\n")
cat("p =", p, "\n")
cat("J =", J, "\n")
```

```{r}
# check
head(X)
table(df_model$month)
table(df_model$day_of_week)
summary(df_model$station_id)
```
# 6. Prior specification

```{r priors-and-initialization}
# -----------------------------
# Prior hyperparameters
# -----------------------------

# beta ~ N(b0, B0)
c_beta <- 100
b0 <- rep(0, p)
B0 <- diag(c_beta, p)
B0_inv <- solve(B0)

# sigma2 ~ IG(a_sigma, b_sigma)
a_sigma <- 2
b_sigma <- 0.5

# tau ~ Half-Cauchy(0, A)
A_tau <- 5

# -----------------------------
# MCMC settings
# -----------------------------
n_iter <- 5000
burn_in <- 1000

# proposal sd for log(tau)
proposal_sd_tau <- 0.15

# -----------------------------
# Initial values
# -----------------------------
beta_curr <- rep(0, p)
alpha_curr <- rep(0, J)
sigma2_curr <- var(y_tilde)
tau_curr <- 1

# -----------------------------
# Storage
# -----------------------------
beta_store <- matrix(NA_real_, nrow = n_iter, ncol = p)
alpha_store <- matrix(NA_real_, nrow = n_iter, ncol = J)
sigma2_store <- numeric(n_iter)
tau_store <- numeric(n_iter)
accept_tau <- logical(n_iter)
```

# 7. Full conditional formulas used in the sampler

The sampler uses the following updates:

\[
\beta \mid \cdot \sim \mathcal N(m_\beta, V_\beta)
\]

with

\[
V_\beta = \left(B_0^{-1} + \frac{1}{\sigma^2}X'X\right)^{-1},
\qquad
m_\beta = V_\beta\left(B_0^{-1}b_0 + \frac{1}{\sigma^2}X'(\tilde y - Z\alpha)\right)
\]

\[
\alpha_j \mid \cdot \sim \mathcal N(m_j, V_j)
\]

with

\[
V_j = \left(\frac{n_j}{\sigma^2} + \frac{1}{\tau^2}\right)^{-1},
\qquad
m_j = V_j\frac{1}{\sigma^2}\sum_{i \in j}(\tilde y_i - x_i'\beta)
\]

\[
\sigma^2 \mid \cdot \sim IG\left(a_\sigma + \frac{n}{2}, b_\sigma + \frac{SSR}{2}\right)
\]

For \(\tau\), the full conditional is only known up to proportionality, so a Metropolis-Hastings step is used.

# 8. Utility functions for the sampler

```{r utility-functions}
# Draw from Inverse-Gamma(shape, scale) under the convention:
# p(x) prop to x^{-(a+1)} exp(-b/x)
# If G ~ Gamma(shape=a, rate=b), then 1/G ~ IG(a,b)
rinvgamma <- function(n, shape, scale) {
  1 / rgamma(n, shape = shape, rate = scale)
}

# Log-target for eta = log(tau)
log_target_eta <- function(eta, alpha, A_tau) {
  J <- length(alpha)
  S_alpha <- sum(alpha^2)
  -(J - 1) * eta - S_alpha / (2 * exp(2 * eta)) - log(1 + exp(2 * eta) / A_tau^2)
}
```

# 9. Metropolis-within-Gibbs sampler

```{r sampler-function}
run_mwg_sampler <- function(y_tilde, X, station_index, n_j,
                            b0, B0, a_sigma, b_sigma, A_tau,
                            n_iter = 5000,
                            proposal_sd = 0.15,
                            seed = 123) {

  set.seed(seed)

  n <- length(y_tilde)
  p <- ncol(X)
  J <- length(n_j)
  B0_inv <- solve(B0)

  if (any(is.na(station_index))) stop("station_index contains NA")
  if (min(station_index) < 1) stop("station_index contains values < 1")
  if (max(station_index) > J) stop("station_index exceeds number of groups")
  if (burn_in >= n_iter) stop("burn_in must be smaller than n_iter")

  # Storage
  beta_store <- matrix(NA_real_, nrow = n_iter, ncol = p)
  alpha_store <- matrix(NA_real_, nrow = n_iter, ncol = J)
  sigma2_store <- numeric(n_iter)
  tau_store <- numeric(n_iter)
  accept_tau <- logical(n_iter)

  # Initial values
  beta_curr <- rep(0, p)
  alpha_curr <- rep(0, J)
  sigma2_curr <- var(y_tilde)
  tau_curr <- 1

  XtX <- crossprod(X)

  for (t in seq_len(n_iter)) {

    # ---- 1. Gibbs update for beta ----
    alpha_vec <- alpha_curr[station_index]

    precision_beta <- B0_inv + XtX / sigma2_curr
    V_beta <- solve(precision_beta)
    V_beta <- 0.5 * (V_beta + t(V_beta))

    term_prior <- B0_inv %*% b0
    term_lik <- crossprod(X, y_tilde - alpha_vec) / sigma2_curr
    rhs_beta <- term_prior + term_lik
    m_beta <- solve(precision_beta, rhs_beta)

    if (any(!is.finite(m_beta))) {
      stop(paste("m_beta contains non-finite values at iteration", t))
    }

    beta_curr <- as.numeric(MASS::mvrnorm(
      1, 
      mu = as.numeric(m_beta), 
      Sigma = V_beta
      ))

    # ---- 2. Gibbs updates for alpha_j ----
    xb <- as.numeric(X %*% beta_curr)

    for (j in seq_len(J)) {
      idx <- which(station_index == j)
      V_j <- 1 / (n_j[j] / sigma2_curr + 1 / (tau_curr^2))
      m_j <- V_j * sum(y_tilde[idx] - xb[idx]) / sigma2_curr
      alpha_curr[j] <- rnorm(1, mean = m_j, sd = sqrt(V_j))
    }

    # ---- 3. Gibbs update for sigma^2 ----
    resid <- y_tilde - xb - alpha_curr[station_index]
    SSR <- sum(resid^2)
    sigma2_curr <- rinvgamma(
      1, 
      shape = a_sigma + n / 2, 
      scale = b_sigma + SSR / 2
      )

    if (!is.finite(sigma2_curr) || sigma2_curr <= 0) {
      stop(paste("sigma2_curr became invalid at iteration", t))
    }

    # ---- 4. MH update for tau on log scale ----
    eta_curr <- log(tau_curr)
    eta_prop <- rnorm(1, mean = eta_curr, sd = proposal_sd)

    log_r <- log_target_eta(eta_prop, alpha_curr, A_tau) -
      log_target_eta(eta_curr, alpha_curr, A_tau)

    if (log(runif(1)) < log_r) {
      tau_curr <- exp(eta_prop)
      accept_tau[t] <- TRUE
    } else {
      accept_tau[t] <- FALSE
    }

    # ---- Store current draws ----
    beta_store[t, ] <- beta_curr
    alpha_store[t, ] <- alpha_curr
    sigma2_store[t] <- sigma2_curr
    tau_store[t] <- tau_curr
  }

  list(
    beta = beta_store,
    alpha = alpha_store,
    sigma2 = sigma2_store,
    tau = tau_store,
    accept_tau = accept_tau,
    accept_rate_tau_all = mean(accept_tau)
  )
}
```

# 9b. Postprocessing
```{r}
post_process_mcmc <- function(mcmc_fit, burn_in = 100, thin = 1) {
  n_iter <- nrow(mcmc_fit$beta)
  if (burn_in >= n_iter) stop("burn_in must be smaller than total iterations")
  if (thin < 1) stop("thin must be >= 1")
  
  kept <- seq(from = burn_in + 1, to = n_iter, by = thin)

  list(
    post_beta = mcmc_fit$beta[kept, , drop = FALSE],
    post_alpha = mcmc_fit$alpha[kept, , drop = FALSE],
    post_sigma2 = mcmc_fit$sigma2[kept],
    post_tau = mcmc_fit$tau[kept],
    accept_rate_tau = mean(mcmc_fit$accept_tau[kept]),
    kept_iterations = kept
  )
}

```

# 10. Run the sampler

```{r run-sampler, cache = TRUE}
# MASS::mvrnorm is used inside the function
library(MASS)

mcmc_fit <- run_mwg_sampler(
  y_tilde = y_tilde,
  X = X,
  station_index = station_index,
  n_j = n_j,
  b0 = b0,
  B0 = B0,
  a_sigma = a_sigma,
  b_sigma = b_sigma,
  A_tau = A_tau,
  n_iter = 5000,
  proposal_sd = 0.2,
  seed = 123
)

mcmc_fit$accept_rate_tau_all
```
# 10b. Trace plots

```{r trace-plots, fig.width=10, fig.height=8}
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))

plot(mcmc_fit$beta[, 1], type = "l",
     main = expression(paste("Trace plot of ", beta[1])),
     xlab = "Iteration", ylab = "Value")

plot(mcmc_fit$beta[, 2], type = "l",
     main = expression(paste("Trace plot of ", beta[2])),
     xlab = "Iteration", ylab = "Value")

plot(mcmc_fit$sigma2, type = "l",
     main = expression(paste("Trace plot of ", sigma^2)),
     xlab = "Iteration", ylab = "Value")

plot(mcmc_fit$tau, type = "l",
     main = expression(paste("Trace plot of ", tau)),
     xlab = "Iteration", ylab = "Value")
```


```{r trace-plots-alpha, fig.width=10, fig.height=8}
J <- ncol(mcmc_fit$alpha)
alpha_idx <- unique(c(1, round(J / 2), J))

par(mfrow = c(length(alpha_idx), 1), mar = c(4, 4, 2, 1))

for (j in alpha_idx) {
  plot(mcmc_fit$alpha[, j], type = "l",
       main = bquote(paste("Trace plot of ", alpha[.(j)])),
       xlab = "Iteration", ylab = "Value")
}
```

# 10c. ACF
```{r acf-plots, fig.width=10, fig.height=8}
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))

acf(mcmc_fit$beta[, 1],
    main = expression(paste("ACF of ", beta[1])))

acf(mcmc_fit$beta[, 2],
    main = expression(paste("ACF of ", beta[2])))

acf(mcmc_fit$sigma2,
    main = expression(paste("ACF of ", sigma^2)))

acf(mcmc_fit$tau,
    main = expression(paste("ACF of ", tau)))
```

```{r acf-plots-alpha, fig.width=10, fig.height=8}
par(mfrow = c(length(alpha_idx), 1), mar = c(4, 4, 2, 1))

for (j in alpha_idx) {
  acf(mcmc_fit$alpha[, j],
      main = bquote(paste("ACF of ", alpha[.(j)])))
}
```

# 10d. Choose burn in iterations
```{r choose-burnin}
burn_in <- 200

post <- post_process_mcmc(mcmc_fit, burn_in = burn_in, thin = 5)

post$accept_rate_tau
```

#10e. Trace plots after burn-in
```{r post-burnin-trace, fig.width=10, fig.height=8}
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))

plot(post$post_beta[, 1], type = "l",
     main = expression(paste("Post-burn-in trace of ", beta[1])),
     xlab = "Iteration kept", ylab = "Value")

plot(post$post_beta[, 2], type = "l",
     main = expression(paste("Post-burn-in trace of ", beta[2])),
     xlab = "Iteration kept", ylab = "Value")

plot(post$post_sigma2, type = "l",
     main = expression(paste("Post-burn-in trace of ", sigma^2)),
     xlab = "Iteration kept", ylab = "Value")

plot(post$post_tau, type = "l",
     main = expression(paste("Post-burn-in trace of ", tau)),
     xlab = "Iteration kept", ylab = "Value")
```
### Visual inspection
The trace plots of the post-burn-in chains show no evidence of non-stationarity or systematic trends. All parameters fluctuate around stable levels, suggesting that the burn-in period has successfully removed the initial transient phase. The chains exhibit satisfactory mixing, although some residual autocorrelation is visible for selected parameters. Overall, the visual inspection supports convergence and justifies the use of the retained samples for posterior inference.

# 10f. ACF post burn-in
```{r post-burnin-acf, fig.width=10, fig.height=8}
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))

acf(post$post_beta[, 1],
    main = expression(paste("Post-burn-in ACF of ", beta[1])))

acf(post$post_beta[, 2],
    main = expression(paste("Post-burn-in ACF of ", beta[2])))

acf(post$post_sigma2,
    main = expression(paste("Post-burn-in ACF of ", sigma^2)))

acf(post$post_tau,
    main = expression(paste("Post-burn-in ACF of ", tau)))
```
# 10g. ESS
```{r}
library(coda)

mcmc_beta   <- mcmc(post$post_beta)
mcmc_alpha  <- mcmc(post$post_alpha)
mcmc_sigma2 <- mcmc(post$post_sigma2)
mcmc_tau    <- mcmc(post$post_tau)

effectiveSize(mcmc_beta)
effectiveSize(mcmc_alpha)
effectiveSize(mcmc_sigma2)
effectiveSize(mcmc_tau)
```


Due to autocorrelation in the MCMC draws, the effective sample size (ESS) is used to assess the amount of independent information in the chain. The ESS values are sufficiently large, indicating good mixing of the sampler.

# 11. Posterior summaries

```{r posterior-summaries}
posterior_summary <- function(x) {
  c(
    mean = mean(x),
    sd = sd(x),
    q2.5 = quantile(x, 0.025),
    median = median(x),
    q97.5 = quantile(x, 0.975)
  )
}

beta_summary <- t(apply(post$post_beta, 2, posterior_summary))
sigma2_summary <- posterior_summary(post$post_sigma2)
tau_summary <- posterior_summary(post$post_tau)

beta_summary
sigma2_summary
tau_summary
```
Inspection of the trace plots suggested only a short transient phase, so the first 200 iterations were discarded as burn-in. Residual autocorrelation remained for some parameters, especially the intercept and selected random effects; therefore, a mild thinning scheme (every 5th draw) was applied. The retained chains appeared stable, and the acceptance rate for the Metropolis-Hastings update of τ was satisfactory.


# 13. Sensitivity analysis skeleton

This section presents a simple sensitivity analysis to compare prior choices such as:
- \(c \in \{10, 100, 1000\}\) for \(B_0 = cI\)
- \(A \in \{2.5, 5, 10\}\) for the half-Cauchy prior on \(\tau\)
- \(b_\sigma \in \{0.5, 1\}\) for the inverse-gamma prior on \(\sigma^2\)

```{r sensitivity-template, cache = TRUE}
run_one_spec <- function(c_prior, A_tau, b_sigma,
                         scenario = NA_character_,
                         n_iter = 5000,
                         burn_in = 200,
                         thin = 5,
                         seed = 123) {

  fit <- run_mwg_sampler(
    y_tilde = y_tilde,
    X = X,
    station_index = station_index,
    n_j = n_j,
    b0 = rep(0, ncol(X)),
    B0 = diag(c_prior, ncol(X)),
    a_sigma = 2,
    b_sigma = b_sigma,
    A_tau = A_tau,
    n_iter = n_iter,
    proposal_sd = 0.2,
    seed = seed
  )

  post <- post_process_mcmc(
    mcmc_fit = fit,
    burn_in = burn_in,
    thin = thin
  )

  data.frame(
    scenario = scenario,
    c_prior = c_prior,
    A_tau = A_tau,
    b_sigma = b_sigma,
    beta_intercept_mean = mean(post$post_beta[, 1]),
    sigma2_mean = mean(post$post_sigma2),
    tau_mean = mean(post$post_tau),
    tau_accept_post = post$accept_rate_tau
  )
}
```

# One-at-a-time sensitivity analysis

```{r sensitivity-runs}
baseline_row <- data.frame(
  scenario = "Baseline",
  c_prior = 100,
  A_tau = 5,
  b_sigma = 0.5,
  beta_intercept_mean = mean(post$post_beta[, 1]),
  sigma2_mean = mean(post$post_sigma2),
  tau_mean = mean(post$post_tau),
  tau_accept_post = post$accept_rate_tau
)

sensitivity_results <- bind_rows(
  baseline_row,
  run_one_spec(10,  5, 0.5, scenario = "B0 smaller"),
  run_one_spec(1000, 5, 0.5, scenario = "B0 larger"),
  run_one_spec(100, 2.5, 0.5, scenario = "A_tau smaller"),
  run_one_spec(100, 10, 0.5, scenario = "A_tau larger"),
  run_one_spec(100, 5, 1, scenario = "b_sigma larger")
)

sensitivity_results
```

The sensitivity analysis shows that posterior estimates are remarkably stable across a wide range of prior specifications. In particular, the posterior mean of the intercept, the residual variance σ^2, and the hyperparameter τ remain virtually unchanged when varying the prior scale c, the hyperparameter A, and the inverse-gamma parameter b_σ. As expected, the hyperparameter τ shows slightly higher sensitivity compared to the regression coefficients, although the magnitude of variation remains negligible.
This suggests that the data are highly informative and that the posterior distribution is largely driven by the likelihood rather than the prior. The results are therefore robust to reasonable prior choices.

# 14. CODA: GEWEKE
```{r geweke-diagnostic, message=FALSE, warning=FALSE}
library(coda)

mcmc_beta1  <- mcmc(post$post_beta[, 1])
mcmc_beta2  <- mcmc(post$post_beta[, 2])
mcmc_sigma2 <- mcmc(post$post_sigma2)
mcmc_tau    <- mcmc(post$post_tau)

geweke_results <- data.frame(
  parameter = c("beta_1", "beta_2", "sigma2", "tau"),
  z_score = c(
    geweke.diag(mcmc_beta1)$z,
    geweke.diag(mcmc_beta2)$z,
    geweke.diag(mcmc_sigma2)$z,
    geweke.diag(mcmc_tau)$z
  )
)

geweke_results
```
```{r geweke-alpha}
J <- ncol(post$post_alpha)
alpha_idx <- unique(c(1, round(J / 2), J))

alpha_geweke <- data.frame(
  parameter = paste0("alpha_", alpha_idx),
  z_score = sapply(alpha_idx, function(j) geweke.diag(mcmc(post$post_alpha[, j]))$z)
)

alpha_geweke
```
The Geweke diagnostic provides no evidence against convergence. All z-scores are well below the conventional threshold of 1.96 in absolute value, both for the main model parameters and for the selected random effects. This finding is consistent with the visual inspection of the trace plots and supports the use of the retained draws for posterior inference.


# 15. PPC
```{r}
# Number of posterior draws to use for PPC
S <- min(100, nrow(post$post_beta))

set.seed(123)

# Randomly select posterior draws
draw_ids <- sample(seq_len(nrow(post$post_beta)), S)

# Matrix to store replicated datasets
yrep_mat <- matrix(NA_real_, nrow = length(y_tilde), ncol = S)

for (s in seq_along(draw_ids)) {
  d <- draw_ids[s]

  beta_s   <- post$post_beta[d, ]
  alpha_s  <- post$post_alpha[d, ]
  sigma2_s <- post$post_sigma2[d]

  mu_s <- as.vector(X %*% beta_s) + alpha_s[station_index]

  yrep_mat[, s] <- rnorm(
    n = length(y_tilde),
    mean = mu_s,
    sd = sqrt(sigma2_s)
  )
}
```

```{r}
obs_mean <- mean(y_tilde)
obs_sd   <- sd(y_tilde)

rep_means <- apply(yrep_mat, 2, mean)
rep_sds   <- apply(yrep_mat, 2, sd)

par(mfrow = c(1, 2))

hist(rep_means, main = "PPC: mean", xlab = "Replicated means")
abline(v = obs_mean, lwd = 2, lty = 2)

hist(rep_sds, main = "PPC: sd", xlab = "Replicated sds")
abline(v = obs_sd, lwd = 2, lty = 2)

par(mfrow = c(1, 1))
```

## Visualization
```{r}
plot(density(y_tilde),
     lwd = 2,
     main = "Posterior Predictive Check",
     xlab = "log(PM2.5)",
     ylim = c(0, max(density(y_tilde)$y) * 1.2))

for (s in 1:S) {
  lines(density(yrep_mat[, s]), col = rgb(1, 0, 0, 0.08))
}

lines(density(y_tilde), lwd = 2)
legend("topright",
       legend = c("Observed data", "Replicated data"),
       lwd = c(2, 1),
       col = c("black", rgb(1, 0, 0, 0.4)),
       bty = "n")
```

