---
title: "Bayesian Hierarchical Model for PM2.5"
author: "Leila Cielok"
date: "29-04-2026"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup-save-figures, include=FALSE}
# directory for paper figures
fig_dir <- "../paper/figures"
if (!dir.exists(fig_dir)) {
  dir.create(fig_dir, recursive = TRUE)
}

# Helper function to save base R plots
save_base_plot <- function(filename, width = 6, height = 4) {
  filepath <- file.path(fig_dir, filename)
  pdf(filepath, width = width, height = height)
  return(filepath)
}

# Helper function to close device
close_plot <- function() dev.off()

```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
set.seed(123)
```

# Introduction

This project implements a Bayesian hierarchical model for daily PM2.5 concentrations observed across multiple monitoring stations. The outcome is modeled on the log scale to account for right-skewness in the raw data.

The specification is:

$$
\tilde y_i = \log(y_i),
\qquad
\tilde y_i \mid \beta, \alpha, \sigma^2 \sim \mathcal N(X_i^\top \beta + \alpha_{j[i]}, \sigma^2)
$$

with station-specific random intercepts

$$
\alpha_j \mid \tau \sim \mathcal N(0, \tau^2),
\qquad j = 1, \dots, J.
$$

A Metropolis-within-Gibbs sampler is used:
- Gibbs updates for $\beta$, $\alpha_j$, and $\sigma^2$
- a Metropolis-Hastings step for $\tau$


# Libraries and data import

```{r libraries}
library(dplyr)
library(ggplot2)
library(readr)
library(e1071)
library(tidyr)
library(stringr)
library(purrr)
library(gridExtra)
```

```{r load-data}
df <- read_csv("../pm25_bayes.csv", show_col_types = FALSE)
str(df)
df$Pollutant <- NULL
df <- df %>%
  mutate(
    Start = as.Date(Start),
    month = factor(month),
    day_of_week = factor(
      day_of_week,
      levels = 0:6,
      labels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")
    )
  )
summary(df)
```
# EDA
## Outcome variable
```{r eda-outcome, message=FALSE, warning=FALSE}
# Basic summary statistics
outcome_var <- "y"

summary(df[[outcome_var]])

# Missing and non-positive values
cat("Number of missing values:", sum(is.na(df[[outcome_var]])), "\n")
cat("Number of non-positive values:", sum(df[[outcome_var]] <= 0, na.rm = TRUE), "\n")

# Histogram of raw outcome
ggplot(df, aes(x = .data[[outcome_var]])) +
  geom_histogram(bins = 40) +
  labs(
    title = "Distribution of the raw PM2.5 outcome",
    x = outcome_var,
    y = "Frequency"
  )

# Skeweness of raw data
skew_raw <- skewness(df[[outcome_var]], na.rm = TRUE)
cat("Skewness (raw):", skew_raw, "\n")

# Histogram of log-transformed outcome
df_pos <- df %>% filter(.data[[outcome_var]] > 0)

p1 <- ggplot(df_pos, aes(x = log(.data[[outcome_var]]))) +
  geom_histogram(bins = 40) +
  labs(
    x = paste0("log(", outcome_var, ")"),
    y = "Frequency"
  )
ggsave(
  filename = file.path(fig_dir, "hist_log_pm25.pdf"),
  plot = p1,
  width = 6,
  height = 4
)

# Skeweness of log data
skew_log <- skewness(log(df[[outcome_var]]), na.rm = TRUE)
cat("Skewness (log):", skew_log, "\n")

# Compare raw and log-scale summaries
cat("\nSummary of raw outcome:\n")
print(summary(df_pos[[outcome_var]]))

cat("\nSummary of log-transformed outcome:\n")
print(summary(log(df_pos[[outcome_var]])))
```

```{r save-combined-outcome}
save_base_plot("hist_raw_log_pm25.pdf", 8, 4.5)

par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))

hist(df[[outcome_var]],
     breaks = 40,
     main = "Raw PM2.5",
     xlab = "PM2.5",
     ylab = "Frequency")

hist(log(df_pos[[outcome_var]]),
     breaks = 40,
     main = "Log-transformed PM2.5",
     xlab = "log(PM2.5)",
     ylab = "Frequency")

par(mfrow = c(1, 1))

close_plot()
```


## Data

```{r eda-panel-structure, message=FALSE, warning=FALSE}
# 1. Observations per station
obs_per_station <- df %>%
  count(station_id, name = "n_obs") %>%
  arrange(desc(n_obs))

cat("Number of stations:", n_distinct(df$station_id), "\n")
cat("Min observations per station:", min(obs_per_station$n_obs), "\n")
cat("Median observations per station:", median(obs_per_station$n_obs), "\n")
cat("Max observations per station:", max(obs_per_station$n_obs), "\n")

head(obs_per_station, 10)

# 2a. Observations per month
obs_per_month <- df %>%
  count(month, name = "n_obs") %>%
  arrange(month)

print(obs_per_month)

# 2b. Observations per day of week
obs_per_day <- df %>%
  count(day_of_week, name = "n_obs")

print(obs_per_day)

# 3. Station-month panel structure
station_month <- df %>%
  distinct(station_id, month) %>%
  count(station_id, name = "n_months_present") %>%
  arrange(n_months_present, station_id)

cat("Min number of months observed per station:", min(station_month$n_months_present), "\n")
cat("Median number of months observed per station:", median(station_month$n_months_present), "\n")
cat("Max number of months observed per station:", max(station_month$n_months_present), "\n")

head(station_month, 10)

# 4. Full station x month table and missing months
station_month_full <- expand_grid(
  station_id = sort(unique(df$station_id)),
  month = sort(unique(df$month))
) %>%
  left_join(
    df %>%
      count(station_id, month, name = "n_obs"),
    by = c("station_id", "month")
  ) %>%
  mutate(
    n_obs = ifelse(is.na(n_obs), 0, n_obs),
    missing_month = ifelse(n_obs == 0, 1, 0)
  )

missing_months_by_station <- station_month_full %>%
  group_by(station_id) %>%
  summarise(
    missing_months = sum(missing_month),
    months_observed = sum(n_obs > 0),
    .groups = "drop"
  ) %>%
  arrange(desc(missing_months), station_id)

cat("Stations with at least one completely missing month:",
    sum(missing_months_by_station$missing_months > 0), "\n")

head(missing_months_by_station, 15)

# 5. Months missing for each station
missing_month_detail <- station_month_full %>%
  filter(n_obs == 0) %>%
  arrange(station_id, month)

head(missing_month_detail, 20)

# ---- Plots for the paper ----

p_obs_station <- ggplot(obs_per_station, aes(x = n_obs)) +
  geom_histogram(
    bins = 24,
    fill = "grey45",
    color = "white",
    linewidth = 0.2
  ) +
  labs(
    x = "Number of observations per station",
    y = "Frequency"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )

p_months_station <- ggplot(station_month, aes(x = n_months_present)) +
  geom_bar(
    fill = "grey45",
    color = "white",
    linewidth = 0.2
  ) +
  scale_x_continuous(breaks = 4:12) +
  labs(
    x = "Number of months observed per station",
    y = "Frequency"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )

p_panel_structure <- gridExtra::arrangeGrob(
  p_obs_station,
  p_months_station,
  ncol = 2
)

# Save figure for the paper
ggsave(
  filename = file.path(fig_dir, "panel_structure_summary.pdf"),
  plot = p_panel_structure,
  width = 7,
  height = 3.8,
  units = "in"
)

grid::grid.newpage()
grid::grid.draw(p_panel_structure)
```

Exploratory analysis shows that the raw PM2.5 outcome is highly right-skewed, with a skewness of 17.06 and a long upper tail driven by extreme observations. After log transformation, skewness drops to -0.29, yielding a much more symmetric distribution and supporting the use of a Gaussian likelihood for the transformed outcome.

The panel structure is markedly unbalanced across monitoring stations. The number of observations per station ranges from 69 to 332, with a median of 299.5. The dataset is also uneven across months: while most months contain between roughly 2,600 and 3,100 observations, January is severely underrepresented, with only 177 observations. Moreover, 32 stations exhibit at least one completely missing month, and the minimum temporal coverage is only 4 months for some stations. These features support the use of a hierarchical specification, which is well suited to unbalanced grouped data.

# Data preprocessing

```{r preprocessing}
# Create log outcome and keep only finite values
df <- df %>%
  mutate(log_y = log(.data[[outcome_var]])) %>%
  filter(is.finite(log_y))

# Keep only variables used in the model and clean them
model_vars <- c("log_y", "station_id", "month", "day_of_week")

df_model <- df %>%
  dplyr::select(all_of(model_vars)) %>%
  drop_na() %>%
  mutate(
    station_id = factor(station_id),
    month = factor(month),
    day_of_week = factor(day_of_week)
  )

cat("Number of observations:", nrow(df_model), "\n")
cat("Number of stations:", n_distinct(df_model$station_id), "\n")

# Summary of transformed outcome
df_model %>%
  summarise(
    n = n(),
    mean_log_y = mean(log_y),
    sd_log_y = sd(log_y),
    min_log_y = min(log_y),
    max_log_y = max(log_y)
  )
```

# Design matrices and indexing

```{r design-matrices}
# Fixed-effects matrix
X <- model.matrix(~ month + day_of_week, data = df_model)

# Outcome vector
y_tilde <- df_model$log_y

# Group index for stations
station_index <- as.integer(df_model$station_id)

# Dimensions
J <- nlevels(df_model$station_id)
n <- length(y_tilde)
p <- ncol(X)

# Number of observations per group
n_j <- as.integer(table(df_model$station_id))

cat("n =", n, "\n")
cat("p =", p, "\n")
cat("J =", J, "\n")
```

```{r}
# check
head(X)
table(df_model$month)
table(df_model$day_of_week)
summary(df_model$station_id)
```
# Prior specifications

Weakly informative priors are adopted for all model parameters, reflecting limited prior knowledge while ensuring posterior regularization.

For the regression coefficients, a multivariate normal prior is assumed:

$$
\beta \sim \mathcal N(b_0, B_0),
$$

where $b_0$ is a prior mean vector and $B_0 = cI$ is a diagonal covariance matrix with $c$ controlling the prior variance.

For the station-specific random effects, I specify:

$$
\alpha_j \mid \tau \sim \mathcal N(0, \tau^2), \qquad j = 1, \dots, J,
$$

which induces partial pooling across monitoring stations.

For the observation noise variance, an inverse-gamma prior is assumed:

$$
\sigma^2 \sim \mathrm{IG}(a_\sigma, b_\sigma),
$$

with small values of $a_\sigma$ and $b_\sigma$ to obtain a weakly informative specification.

Finally, for the scale parameter of the random effects, a half-Cauchy prior is used:

$$
\tau \sim \mathrm{C}^+(0, A),
$$

which is commonly recommended for hierarchical variance components due to its heavy tails and robustness properties.

These choices are motivated by both modeling and computational considerations. 
In particular, conjugate priors are adopted for $\beta$, $\alpha_j$, and $\sigma^2$ to allow for efficient Gibbs updates, while a non-conjugate prior is specified for $\tau$, requiring a Metropolis–Hastings step. 
Overall, the specification ensures sufficient flexibility while preventing overfitting, and will be further assessed through a prior sensitivity analysis.

```{r priors-and-initialization}
# -----------------------------
# Prior hyperparameters
# -----------------------------

# beta ~ N(b0, B0)
c_beta <- 100
b0 <- rep(0, p)
B0 <- diag(c_beta, p)
B0_inv <- solve(B0)

# sigma2 ~ IG(a_sigma, b_sigma)
a_sigma <- 2
b_sigma <- 0.5

# tau ~ Half-Cauchy(0, A)
A_tau <- 5

# -----------------------------
# MCMC settings
# -----------------------------
n_iter <- 10000
burn_in <- 1000

# proposal sd for log(tau)
proposal_sd_tau <- 0.2

# -----------------------------
# Initial values
# -----------------------------
beta_curr <- rep(0, p)
alpha_curr <- rep(0, J)
sigma2_curr <- var(y_tilde)
tau_curr <- 1

# -----------------------------
# Storage
# -----------------------------
beta_store <- matrix(NA_real_, nrow = n_iter, ncol = p)
alpha_store <- matrix(NA_real_, nrow = n_iter, ncol = J)
sigma2_store <- numeric(n_iter)
tau_store <- numeric(n_iter)
accept_tau <- logical(n_iter)
```

# Full conditional conditional posterior distributions specification

The full conditional posterior distributions are derived and passed to the Gibbs sampling algorithm:

$$
\beta \mid \cdot \sim \mathcal N(m_\beta, V_\beta)
$$

with

$$
V_\beta = \left(B_0^{-1} + \frac{1}{\sigma^2}X^\top X\right)^{-1},
\qquad
m_\beta = V_\beta\left(B_0^{-1}b_0 + \frac{1}{\sigma^2}X^\top(\tilde y - Z\alpha)\right)
$$

$$
\alpha_j \mid \cdot \sim \mathcal N(m_j, V_j)
$$

with

$$
V_j = \left(\frac{n_j}{\sigma^2} + \frac{1}{\tau^2}\right)^{-1},
\qquad
m_j = V_j \frac{1}{\sigma^2} \sum_{i \in j} (\tilde y_i - x_i^\top \beta)
$$

$$
\sigma^2 \mid \cdot \sim \mathrm{IG}\left(a_\sigma + \frac{n}{2},\; b_\sigma + \frac{SSR}{2}\right)
$$

where

$$
SSR = \sum_{i=1}^n (\tilde y_i - x_i^\top \beta - \alpha_{j[i]})^2.
$$

The full conditional distribution of $\tau$ is not available in closed form. 
Up to proportionality, it is given by:

$$
p(\tau \mid \alpha) \propto \left[ \prod_{j=1}^J \mathcal N(\alpha_j \mid 0, \tau^2) \right] \cdot p(\tau)
$$

Assuming a half-Cauchy prior $\tau \sim \text{C}^+(0, A)$, this yields:

$$
p(\tau \mid \alpha) \propto \tau^{-J} \exp\left(-\frac{1}{2\tau^2} \sum_{j=1}^J \alpha_j^2 \right) \cdot \frac{1}{1 + (\tau/A)^2}
$$

This distribution does not belong to a standard family, hence we use a Metropolis–Hastings step within the Gibbs sampler.


# Metropolis-within-Gibbs sampler

```{r sampler-function}
run_mwg_sampler <- function(y_tilde, X, station_index, n_j,
                            b0, B0, a_sigma, b_sigma, A_tau,
                            n_iter = 10000,
                            proposal_sd = 0.2,
                            seed = 123) {

  set.seed(seed)

  n <- length(y_tilde)
  p <- ncol(X)
  J <- length(n_j)
  B0_inv <- solve(B0)

  if (any(is.na(station_index))) stop("station_index contains NA")
  if (min(station_index) < 1) stop("station_index contains values < 1")
  if (max(station_index) > J) stop("station_index exceeds number of groups")

  # Storage
  beta_store <- matrix(NA_real_, nrow = n_iter, ncol = p)
  alpha_store <- matrix(NA_real_, nrow = n_iter, ncol = J)
  sigma2_store <- numeric(n_iter)
  tau_store <- numeric(n_iter)
  accept_tau <- logical(n_iter)

  # Initial values
  beta_curr <- rep(0, p)
  alpha_curr <- rep(0, J)
  sigma2_curr <- var(y_tilde)
  tau_curr <- 1

  XtX <- crossprod(X)

  for (t in seq_len(n_iter)) {

    # ---- 1. Gibbs update for beta ----
    alpha_vec <- alpha_curr[station_index]

    precision_beta <- B0_inv + XtX / sigma2_curr
    V_beta <- solve(precision_beta)
    V_beta <- 0.5 * (V_beta + t(V_beta))

    term_prior <- B0_inv %*% b0
    term_lik <- crossprod(X, y_tilde - alpha_vec) / sigma2_curr
    rhs_beta <- term_prior + term_lik
    m_beta <- solve(precision_beta, rhs_beta)

    if (any(!is.finite(m_beta))) {
      stop(paste("m_beta contains non-finite values at iteration", t))
    }

    beta_curr <- as.numeric(MASS::mvrnorm(
      1, 
      mu = as.numeric(m_beta), 
      Sigma = V_beta
      ))

    # ---- 2. Gibbs updates for alpha_j ----
    xb <- as.numeric(X %*% beta_curr)

    for (j in seq_len(J)) {
      idx <- which(station_index == j)
      V_j <- 1 / (n_j[j] / sigma2_curr + 1 / (tau_curr^2))
      m_j <- V_j * sum(y_tilde[idx] - xb[idx]) / sigma2_curr
      alpha_curr[j] <- rnorm(1, mean = m_j, sd = sqrt(V_j))
    }

    # ---- 3. Gibbs update for sigma^2 ----
    resid <- y_tilde - xb - alpha_curr[station_index]
    SSR <- sum(resid^2)
    sigma2_curr <- rinvgamma(
      1, 
      shape = a_sigma + n / 2, 
      scale = b_sigma + SSR / 2
      )

    if (!is.finite(sigma2_curr) || sigma2_curr <= 0) {
      stop(paste("sigma2_curr became invalid at iteration", t))
    }

    # ---- 4. MH update for tau on log scale ----
    eta_curr <- log(tau_curr)
    eta_prop <- rnorm(1, mean = eta_curr, sd = proposal_sd)

    log_r <- log_target_eta(eta_prop, alpha_curr, A_tau) -
      log_target_eta(eta_curr, alpha_curr, A_tau)

    if (log(runif(1)) < log_r) {
      tau_curr <- exp(eta_prop)
      accept_tau[t] <- TRUE
    } else {
      accept_tau[t] <- FALSE
    }

    # ---- Store current draws ----
    beta_store[t, ] <- beta_curr
    alpha_store[t, ] <- alpha_curr
    sigma2_store[t] <- sigma2_curr
    tau_store[t] <- tau_curr
  }

  list(
    beta = beta_store,
    alpha = alpha_store,
    sigma2 = sigma2_store,
    tau = tau_store,
    accept_tau = accept_tau,
    accept_rate_tau_all = mean(accept_tau)
  )
}
```

## Utility functions for the sampler

```{r utility-functions}
# Draw from Inverse-Gamma(shape, scale) under the convention:
# p(x) prop to x^{-(a+1)} exp(-b/x)
# If G ~ Gamma(shape=a, rate=b), then 1/G ~ IG(a,b)
rinvgamma <- function(n, shape, scale) {
  1 / rgamma(n, shape = shape, rate = scale)
}

# Log-target for eta = log(tau)
log_target_eta <- function(eta, alpha, A_tau) {
  J <- length(alpha)
  S_alpha <- sum(alpha^2)
  -(J - 1) * eta - S_alpha / (2 * exp(2 * eta)) - log(1 + exp(2 * eta) / A_tau^2)
}
```

## Postprocessing formula

```{r}
post_process_mcmc <- function(mcmc_fit, burn_in = 500, thin = 1) {
  n_iter <- nrow(mcmc_fit$beta)
  if (burn_in >= n_iter) stop("burn_in must be smaller than total iterations")
  if (thin < 1) stop("thin must be >= 1")
  
  kept <- seq(from = burn_in + 1, to = n_iter, by = thin)

  list(
    post_beta = mcmc_fit$beta[kept, , drop = FALSE],
    post_alpha = mcmc_fit$alpha[kept, , drop = FALSE],
    post_sigma2 = mcmc_fit$sigma2[kept],
    post_tau = mcmc_fit$tau[kept],
    accept_rate_tau = mean(mcmc_fit$accept_tau[kept]),
    kept_iterations = kept
  )
}

```

# Run the sampler

```{r run-sampler, cache = TRUE}
# MASS::mvrnorm is used inside the function
library(MASS)

mcmc_fit <- run_mwg_sampler(
  y_tilde = y_tilde,
  X = X,
  station_index = station_index,
  n_j = n_j,
  b0 = b0,
  B0 = B0,
  a_sigma = a_sigma,
  b_sigma = b_sigma,
  A_tau = A_tau,
  n_iter = 10000,
  proposal_sd = 0.2,
  seed = 123
)

mcmc_fit$accept_rate_tau_all
```

## Visualization: Trace plots

```{r trace-plots, fig.width=10, fig.height=8}
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))

plot(mcmc_fit$beta[, 1], type = "l",
     main = expression(paste("Trace plot of ", beta[1])),
     xlab = "Iteration", ylab = "Value")

plot(mcmc_fit$beta[, 2], type = "l",
     main = expression(paste("Trace plot of ", beta[2])),
     xlab = "Iteration", ylab = "Value")

plot(mcmc_fit$sigma2, type = "l",
     main = expression(paste("Trace plot of ", sigma^2)),
     xlab = "Iteration", ylab = "Value")

plot(mcmc_fit$tau, type = "l",
     main = expression(paste("Trace plot of ", tau)),
     xlab = "Iteration", ylab = "Value")
```


```{r trace-plots-alpha, fig.width=10, fig.height=8}
J <- ncol(mcmc_fit$alpha)
alpha_idx <- unique(c(1, round(J / 2), J))

par(mfrow = c(length(alpha_idx), 1), mar = c(4, 4, 2, 1))

for (j in alpha_idx) {
  plot(mcmc_fit$alpha[, j], type = "l",
       main = bquote(paste("Trace plot of ", alpha[.(j)])),
       xlab = "Iteration", ylab = "Value")
}
```

## Visualization: ACF
```{r acf-plots, fig.width=10, fig.height=8}
par(mfrow = c(2, 2), mar = c(4, 4, 5, 1))

acf(mcmc_fit$beta[, 1],
    main = expression(paste("ACF of ", beta[1])))

acf(mcmc_fit$beta[, 2],
    main = expression(paste("ACF of ", beta[2])))

acf(mcmc_fit$sigma2,
    main = expression(paste("ACF of ", sigma^2)))

acf(mcmc_fit$tau,
    main = expression(paste("ACF of ", tau)))
```

```{r acf-plots-alpha, fig.width=10, fig.height=8}
par(mfrow = c(length(alpha_idx), 1), mar = c(4, 4, 2, 1))

for (j in alpha_idx) {
  acf(mcmc_fit$alpha[, j],
      main = bquote(paste("ACF of ", alpha[.(j)])))
}
```

## Heidelberger-Welch diagnostic for burn-in assessment

```{r heidelberg-diagnostic, warning=FALSE, message=FALSE}
library(coda)

mcmc_full <- mcmc(
  cbind(
    mcmc_fit$beta,
    mcmc_fit$alpha,
    sigma2 = mcmc_fit$sigma2,
    tau = mcmc_fit$tau
  )
)

colnames(mcmc_full) <- c(
  paste0("beta_", seq_len(ncol(mcmc_fit$beta)) - 1),
  paste0("alpha_", seq_len(ncol(mcmc_fit$alpha))),
  "sigma2",
  "tau"
)

heidel_list <- lapply(seq_len(ncol(mcmc_full)), function(k) {
  out <- heidel.diag(mcmc_full[, k])
  
  mat <- unclass(out)
  df <- as.data.frame(mat)
  
  df$Parameter <- colnames(mcmc_full)[k]
  df
})

heidel_df <- do.call(rbind, heidel_list)
rownames(heidel_df) <- NULL

# Suggested burn-in according to the stationarity test
heidel_burnin <- max(heidel_df$start, na.rm = TRUE) - 1

# Number of parameters passing the stationarity and half-width tests
n_parameters <- nrow(heidel_df)
n_pass_stationarity <- sum(heidel_df$stest == 1, na.rm = TRUE)
n_pass_halfwidth <- sum(heidel_df$htest == 1, na.rm = TRUE)

# Parameters failing the stationarity test
heidel_failed_stationarity <- heidel_df[heidel_df$stest == 0, ]

# Parameters failing the half-width test
heidel_failed_halfwidth <- heidel_df[heidel_df$htest == 0, ]

heidel_burnin
n_pass_stationarity
n_pass_halfwidth
n_parameters
heidel_failed_stationarity
heidel_failed_halfwidth

saveRDS(heidel_df, file = file.path(fig_dir, "heidelberg_summary.rds"))
```

## Choose burn in iterations

```{r choose-burnin}
# Burn-in chosen using the Heidelberger-Welch diagnostic.
# A minimum of 500 iterations is retained as a conservative lower bound.
burn_in <- max(500, heidel_burnin)

post <- post_process_mcmc(mcmc_fit, burn_in = burn_in, thin = 4)

saveRDS(post, file = file.path(fig_dir, "post_samples.rds"))

burn_in
post$accept_rate_tau
```

## Trace plots post-burn in
```{r post-burnin-trace, fig.width=6, fig.height=3.5}
save_base_plot("trace_postburn.pdf", 6, 3.5)

par(mfrow = c(1, 2), mar = c(4, 4, 4, 1))

plot(post$post_sigma2, type = "l",
     main = expression(paste("Trace of ", sigma^2)),
     ylab = expression(sigma^2),
     xlab = "Iteration")

plot(post$post_tau, type = "l",
     main = expression(paste("Trace of ", tau)),
     ylab = expression(tau),
     xlab = "Iteration")

par(mfrow = c(1, 1))

close_plot()
```

## ESS 

```{r}
library(coda)

post <- readRDS("../paper/figures/post_samples.rds")

mcmc_beta   <- mcmc(post$post_beta)
mcmc_alpha  <- mcmc(post$post_alpha)
mcmc_sigma2 <- mcmc(post$post_sigma2)
mcmc_tau    <- mcmc(post$post_tau)

ess_beta   <- effectiveSize(mcmc_beta)
ess_alpha  <- effectiveSize(mcmc_alpha)
ess_sigma2 <- effectiveSize(mcmc_sigma2)
ess_tau    <- effectiveSize(mcmc_tau)

ess_all <- data.frame(
  Parameter = c(
    paste0("beta_", seq_along(ess_beta) - 1),
    paste0("alpha_", seq_along(ess_alpha)),
    "sigma2",
    "tau"
  ),
  ESS = c(
    as.numeric(ess_beta),
    as.numeric(ess_alpha),
    as.numeric(ess_sigma2),
    as.numeric(ess_tau)
  )
)

ess_all
summary(ess_all$ESS)
range(ess_all$ESS)
ess_all[order(ess_all$ESS), ][1:10, ]
```


## Visual Inspection

```{r acf-beta-alpha-representative, message=FALSE, warning=FALSE}
library(coda)

# Representative alphas: lowest, median, highest ESS
alpha_idx <- c(
  which.min(ess_alpha),
  order(ess_alpha)[round(length(ess_alpha) / 2)],
  which.max(ess_alpha)
)

# Representative betas: lowest, median, highest ESS
beta_idx <- c(
  which.min(ess_beta),
  order(ess_beta)[round(length(ess_beta) / 2)],
  which.max(ess_beta)
)

alpha_idx
ess_alpha[alpha_idx]

beta_idx
ess_beta[beta_idx]

# ACF of representative alphas
par(mfrow = c(1, 3), mar = c(4, 4, 4, 1))

for (j in alpha_idx) {
  acf(
    post$post_alpha[, j],
    main = bquote("ACF of " ~ alpha[.(j)]),
    xlab = "Lag"
  )
}

par(mfrow = c(1, 1))

# ACF of representative betas
par(mfrow = c(1, 3), mar = c(4, 4, 4, 1))

for (j in beta_idx) {
  acf(
    post$post_beta[, j],
    main = bquote("ACF of " ~ beta[.(j - 1)]),
    xlab = "Lag"
  )
}

par(mfrow = c(1, 1))
```

## ACF post burn-in (sigma and tau)
```{r post-burnin-acf, fig.width=8, fig.height=4}
save_base_plot("acf_postburn.pdf", 7, 3.5)

par(mfrow = c(1, 2), mar = c(4, 4, 4, 1))

acf(post$post_sigma2,
    main = expression(paste("ACF of ", sigma^2)))

acf(post$post_tau,
    main = expression(paste("ACF of ", tau)))

par(mfrow = c(1, 1))

close_plot()
```

# Posterior summaries

```{r posterior-summaries}
posterior_summary <- function(x) {
  c(
    mean = mean(x),
    sd = sd(x),
    q2.5 = quantile(x, 0.025),
    median = median(x),
    q97.5 = quantile(x, 0.975)
  )
}

beta_summary <- t(apply(post$post_beta, 2, posterior_summary))
sigma2_summary <- posterior_summary(post$post_sigma2)
tau_summary <- posterior_summary(post$post_tau)

beta_summary
sigma2_summary
tau_summary
```

## HDI

```{r}
library(coda)
beta_hpd <- HPDinterval(mcmc_beta, prob = 0.95)
sigma2_hpd <- HPDinterval(mcmc_sigma2, prob = 0.95)
tau_hpd    <- HPDinterval(mcmc_tau, prob = 0.95)
```
```{r}
HPDinterval(mcmc_beta)
```

# Sensitivity analysis

This section presents a simple sensitivity analysis to assess the robustness of posterior inference to prior choices. In particular, I consider:

- $c \in \{10, 100, 1000\}$ for the prior covariance $B_0 = cI$
- $A \in \{2.5, 5, 10\}$ for the half-Cauchy prior on $\tau$
- $b_\sigma \in \{0.5, 1\}$ for the inverse-gamma prior on $\sigma^2$

```{r sensitivity-template, cache = TRUE}
run_one_spec <- function(c_prior, A_tau, b_sigma,
                         scenario = NA_character_,
                         n_iter = 5000,
                         burn_in = 200,
                         thin = 5,
                         seed = 123) {

  fit <- run_mwg_sampler(
    y_tilde = y_tilde,
    X = X,
    station_index = station_index,
    n_j = n_j,
    b0 = rep(0, ncol(X)),
    B0 = diag(c_prior, ncol(X)),
    a_sigma = 2,
    b_sigma = b_sigma,
    A_tau = A_tau,
    n_iter = n_iter,
    proposal_sd = 0.2,
    seed = seed
  )

  post <- post_process_mcmc(
    mcmc_fit = fit,
    burn_in = burn_in,
    thin = thin
  )

  data.frame(
    scenario = scenario,
    c_prior = c_prior,
    A_tau = A_tau,
    b_sigma = b_sigma,
    beta_intercept_mean = mean(post$post_beta[, 1]),
    sigma2_mean = mean(post$post_sigma2),
    tau_mean = mean(post$post_tau),
    tau_accept_post = post$accept_rate_tau
  )
}
```

## One-at-a-time sensitivity analysis

```{r sensitivity-runs}
post <- readRDS("../paper/figures/post_samples.rds")

baseline_row <- data.frame(
  scenario = "Baseline",
  c_prior = 100,
  A_tau = 5,
  b_sigma = 0.5,
  beta_intercept_mean = mean(post$post_beta[, 1]),
  sigma2_mean = mean(post$post_sigma2),
  tau_mean = mean(post$post_tau),
  tau_accept_post = post$accept_rate_tau
)

```

```{r}
library(dplyr)


sensitivity_results <- bind_rows(
  baseline_row,
  run_one_spec(10,   5, 0.5, scenario = "Smaller prior variance for $\\beta$"),
  run_one_spec(1000, 5, 0.5, scenario = "Larger prior variance for $\\beta$"),
  run_one_spec(100, 2.5, 0.5, scenario = "Smaller $A_{\\tau}$"),
  run_one_spec(100, 10, 0.5, scenario = "Larger $A_{\\tau}$"),
  run_one_spec(100, 5, 1, scenario = "Larger $b_{\\sigma}$")
)

sensitivity_results

saveRDS(sensitivity_results, file = file.path(fig_dir, "sensitivity_results.rds"))
```


```{r tab-sensitivity, echo=FALSE, warning=FALSE, message=FALSE, results='asis'}
library(dplyr)
library(knitr)

sensitivity_results %>%
  mutate(
    scenario = recode(
      scenario,
      "Baseline" = "Baseline",
      "Smaller prior variance for beta" = "Smaller prior variance for $\\beta$",
      "Larger prior variance for beta" = "Larger prior variance for $\\beta$",
      "Smaller A_tau" = "Smaller $A_{\\tau}$",
      "Larger A_tau" = "Larger $A_{\\tau}$",
      "Larger b_sigma" = "Larger $b_{\\sigma}$"
    ),
    beta_intercept_mean = round(beta_intercept_mean, 3),
    sigma2_mean = round(sigma2_mean, 3),
    tau_mean = round(tau_mean, 3),
    tau_accept_post = round(tau_accept_post, 3),
    c_prior = paste0("$", c_prior, "$"),
    A_tau = paste0("$", A_tau, "$"),
    b_sigma = paste0("$", b_sigma, "$")
  ) %>%
  rename(
    Scenario = scenario,
    `$c$` = c_prior,
    `$A_{\\tau}$` = A_tau,
    `$b_{\\sigma}$` = b_sigma,
    `$\\beta_0$` = beta_intercept_mean,
    `$\\sigma^2$` = sigma2_mean,
    `$\\tau$` = tau_mean,
    `Acceptance ($\\tau$)` = tau_accept_post
  ) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    escape = FALSE,
    caption = "(\\#tab:sensitivity) Sensitivity analysis under alternative prior specifications.",
    align = "lccccccc"
  )
```

The sensitivity analysis shows that posterior estimates are remarkably stable across a wide range of prior specifications. In particular, the posterior mean of the intercept, the residual variance $\sigma^2$, and the hyperparameter $\tau$ remain virtually unchanged when varying the prior scale c, the hyperparameter A, and the inverse-gamma parameter $b_\sigma$. As expected, the hyperparameter $\tau$ shows slightly higher sensitivity compared to the regression coefficients, although the magnitude of variation remains negligible.
This suggests that the data are highly informative and that the posterior distribution is largely driven by the likelihood rather than the prior. The results are therefore robust to reasonable prior choices.

# CODA: Geweke Diagnostics

```{r geweke-diagnostic, message=FALSE, warning=FALSE}
library(coda)
library(dplyr)

geweke_beta <- apply(post$post_beta, 2, function(x) {
  as.numeric(geweke.diag(mcmc(x))$z)
})

geweke_sigma2 <- as.numeric(geweke.diag(mcmc(post$post_sigma2))$z)
geweke_tau    <- as.numeric(geweke.diag(mcmc(post$post_tau))$z)

geweke_alpha <- apply(post$post_alpha, 2, function(x) {
  as.numeric(geweke.diag(mcmc(x))$z)
})

geweke_summary <- data.frame(
  Parameter_block = c("$\\beta$", "$\\alpha$", "$\\sigma^2$", "$\\tau$"),
  Min = c(min(geweke_beta), min(geweke_alpha), geweke_sigma2, geweke_tau),
  Mean = c(mean(geweke_beta), mean(geweke_alpha), geweke_sigma2, geweke_tau),
  Max = c(max(geweke_beta), max(geweke_alpha), geweke_sigma2, geweke_tau)
) %>%
  mutate(across(c(Min, Mean, Max), ~ round(.x, 3)))

saveRDS(geweke_beta, file = file.path(fig_dir, "geweke_beta.rds"))
saveRDS(geweke_alpha, file = file.path(fig_dir, "geweke_alpha.rds"))
saveRDS(geweke_summary, file = file.path(fig_dir, "geweke_summary.rds"))

geweke_summary

```

```{r}
geweke_beta <- geweke.diag(mcmc_beta)$z
geweke_alpha <- geweke.diag(mcmc_alpha)$z

sum(abs(geweke_beta) > 2)
sum(abs(geweke_alpha) > 2)
```


# Posterior Predictive Checks

```{r PPC}
# Number of posterior draws to use for PPC
S <- min(100, nrow(post$post_beta))

set.seed(123)

# Randomly select posterior draws
draw_ids <- sample(seq_len(nrow(post$post_beta)), S)

# Matrix to store replicated datasets
yrep_mat <- matrix(NA_real_, nrow = length(y_tilde), ncol = S)

for (s in seq_along(draw_ids)) {
  d <- draw_ids[s]

  beta_s   <- post$post_beta[d, ]
  alpha_s  <- post$post_alpha[d, ]
  sigma2_s <- post$post_sigma2[d]

  mu_s <- as.vector(X %*% beta_s) + alpha_s[station_index]

  yrep_mat[, s] <- rnorm(
    n = length(y_tilde),
    mean = mu_s,
    sd = sqrt(sigma2_s)
  )
}
```

```{r PPC-mean_sd}
obs_mean <- mean(y_tilde)
obs_sd   <- sd(y_tilde)

rep_means <- apply(yrep_mat, 2, mean)
rep_sds   <- apply(yrep_mat, 2, sd)

save_base_plot("ppc_moments.pdf", 7, 4)

par(mfrow = c(1, 2))

hist(rep_means, main = "PPC: mean")
abline(v = obs_mean, lwd = 2, lty = 2)

hist(rep_sds, main = "PPC: sd")
abline(v = obs_sd, lwd = 2, lty = 2)

par(mfrow = c(1, 1))

close_plot()
```

```{r PPC-plot}
save_base_plot("ppc_density.pdf", 7, 5)

plot(density(y_tilde),
     lwd = 2,
     main = "Posterior Predictive Check",
     xlab = "log(PM2.5)")

for (s in 1:S) {
  lines(density(yrep_mat[, s]), col = rgb(1, 0, 0, 0.08))
}

lines(density(y_tilde), lwd = 2)

legend("topright",
       legend = c("Observed", "Replicated"),
       col = c("black", rgb(1,0,0,0.4)),
       lwd = c(2,1),
       bty = "n")

close_plot()
```
The posterior predictive checks indicate a satisfactory model fit. 
The distribution of replicated means and standard deviations closely matches the observed values, suggesting that the model accurately captures the first two moments of the data. 

Furthermore, the posterior predictive density aligns well with the empirical distribution of $\log(\text{PM2.5})$, with only minor discrepancies in the peak and tail regions. 
Overall, these results suggest that the model provides an adequate representation of the data-generating process, with no major signs of systematic misfit.

# Spatial Analysis
```{r}
library(dplyr)

stations_geo <- df %>%
  distinct(station_id, Municipality, Latitude, Longitude, Altitude)
```

## Average random effects
```{r}
alpha_mean <- colMeans(post$post_alpha)

station_re <- tibble(
  station_id = levels(df_model$station_id),
  alpha_mean = alpha_mean
)

stations_geo <- df %>%
  distinct(station_id, Municipality, Latitude, Longitude, Altitude) %>%
  left_join(station_re, by = "station_id")
```

# Map
```{r}
library(ggplot2)
library(maps)

italy_map <- map_data("world") %>%
  dplyr::filter(region == "Italy")

p_map_alpha <- ggplot() +
  geom_polygon(
    data = italy_map,
    aes(x = long, y = lat, group = group),
    fill = "grey95", color = "grey60", linewidth = 0.2
  ) +
  geom_point(
    data = stations_geo,
    aes(x = Longitude, y = Latitude, color = alpha_mean),
    size = 2.2, alpha = 0.9
  ) +
  # LINEE MACRO-AREA
  geom_hline(yintercept = 44, linetype = "dashed", color = "black", linewidth = 0.4) +
  geom_hline(yintercept = 41.5, linetype = "dashed", color = "black", linewidth = 0.4) +
  coord_quickmap(xlim = c(6, 19), ylim = c(36, 47.5)) +
  labs(
    x = "Longitude",
    y = "Latitude",
    color = expression(hat(alpha)[j])
  ) +
  
  theme_minimal()

filepath <- file.path(fig_dir, "station_random_effects_map.pdf")
ggsave(filepath, plot = p_map_alpha, width = 6.5, height = 5)

```

```{r}
summary(stations_geo$alpha_mean)
sum(is.na(stations_geo$Latitude))
sum(is.na(stations_geo$Longitude))
head(stations_geo)
```

```{r nord-vs-south}
stations_geo <- stations_geo %>%
  mutate(
    macro_area = case_when(
      Latitude >= 44 ~ "North",
      Latitude >= 41.5 ~ "Center",
      TRUE ~ "South"
    )
  )

stations_geo %>%
  group_by(macro_area) %>%
  summarise(
    n = n(),
    mean_alpha = mean(alpha_mean),
    median_alpha = median(alpha_mean),
    sd_alpha = sd(alpha_mean),
    .groups = "drop"
  )
```
```{r nort-vs-sout_bayesian}
station_info <- stations_geo %>%
  arrange(match(station_id, levels(df_model$station_id)))

north_idx  <- which(station_info$macro_area == "North")
center_idx <- which(station_info$macro_area == "Center")
south_idx  <- which(station_info$macro_area == "South")

north_draws  <- rowMeans(post$post_alpha[, north_idx, drop = FALSE])
center_draws <- rowMeans(post$post_alpha[, center_idx, drop = FALSE])
south_draws  <- rowMeans(post$post_alpha[, south_idx, drop = FALSE])

diff_nc <- north_draws - center_draws
diff_ns <- north_draws - south_draws

mean(diff_ns > 0)
```


```{r}
library(coda)
library(dplyr)
library(tibble)

area_draws <- list(
  North = north_draws,
  Center = center_draws,
  South = south_draws,
  `North - South` = diff_ns,
  `North - Center` = diff_nc
)

spatial_table <- lapply(names(area_draws), function(x) {
  draws <- area_draws[[x]]
  hpd <- HPDinterval(as.mcmc(draws), prob = 0.95)
  
  tibble(
    Area = x,
    Mean = mean(draws),
    `95% HPD lower` = hpd[1, "lower"],
    `95% HPD upper` = hpd[1, "upper"],
    `Posterior probability` = mean(draws > 0)
  )
}) %>%
  bind_rows() %>%
  dplyr::mutate(
    Mean = round(Mean, 3),
    `95% HPD interval` = paste0(
      "[", round(`95% HPD lower`, 3),
      ", ", round(`95% HPD upper`, 3), "]"
    ),
    `Pr(>0)` = round(`Posterior probability`, 3)
  ) %>%
  dplyr::select(Area, Mean, `95% HPD interval`, `Posterior probability`)

spatial_table
```
```{r}
saveRDS(spatial_table, file = file.path(fig_dir, "spatial_macroarea_table.rds"))
```


The distribution of posterior station-level random effects reveals substantial heterogeneity across monitoring locations, but a clear positive difference between Northern and Southern Italy. This indicates that geographical variation plays a major role in shaping air pollution patterns, even after controlling for temporal effects.

